### 2020-01-24 paper/sex_selection_spacing-ver3.tex
Thighten up introduction

% Demography abstract
\noindent 
Strong son preference is typically associated with shorter birth spacing in the absence of 
sons, but access to sex selection has the potential to reverse this pattern because each 
abortion extends spacing by six to twelve months. 
I introduce an empirical method that simultaneously accounts for how sex selection 
increases both birth spacing and the likelihood of a son. 
Using the four rounds of India's National Family and Health Surveys, I show that birth 
intervals increased substantially over the last four decades, with the most significant
increases among women most likely to use sex selection.
Specifically, well-educated women with no boys now have significantly longer birth
intervals, and more male-biased sex ratios, than similar women with boys. 
Women with no education still follow the standard pattern of short average spacing when 
they have girls and little evidence of sex selection, with medium-educated women showing 
mixed results.
Furthermore, the traditional pattern of longer average spacing with sons appears to arise 
not from fewer very short birth intervals, but rather from the upper end of the distribution.

@@

% [Why India?]
India has a long history of pro-male preference \citep{Kishor1993,murthi95,arnold98}.
Mortality risk is higher for females than males for most ages groups, resulting in
an almost continuous increase in the ratio of males to females over the last century
 \citep{dyson01,Navaneetham2011,Bongaarts2015}.
Differential stopping behavior is widespread, with an additional child more likely the
fewer sons a family has 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.
The duration to the next birth is also shorter with fewer sons, likely contributing
to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhargava2003,Rutstein2005,Bhalotra2008,Maitra2008,Makepeace2008,
Jayachandran2011,Kumar2016,Jayachandran2017a,Ghosh2018}.
Furthermore, India's total fertility rate has declined substantially and is now below 
replacement in some areas [TK more precision] 
\citep{Guilmoto2013,Dharmalingam2014,International-Institute-for-Population-Sciences-IIPS2017}.
The combination of falling fertility, strong son preference, and expanding access to 
prenatal sex determination dramaticly increased the males-to-females ratio at 
birth over the last three decades
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2009a,Guilmoto2012,
Bongaarts2013,Portner2015b,Jayachandran2017}.

Highlights:
- changes in spacing overall
- much longer spacing with girls
- biases in TFR
- effects on mortality

Birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination potentially 
changed the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion increases the interval between births by six to twelve 
months.%
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Hence, longer birth intervals in the absence of sons, which previously would be taken as 
an indicator of lower son preference, may now instead arise because couples with a strong 
son preference use sex selection.

Here, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
I introduce an empirical method that directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son. 
The method can be used to analyze both situations with and without prenatal
sex selection.
I apply the method to birth histories of Hindu women, using data from the four
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016.

% [Why spacing?] 
There are four main reasons for examining birth spacing and how it changes with 
access to sex selection.
First, researchers have made extensive use of birth spacing as a measure of son preference, 
and it is critical to understand to what extent spacing is still a useful measure of son 
preference.
For example, in Bangladesh, India, and Vietnam, having more boys, and particularly having 
a boy as the last-born child, significantly increase the duration to next birth,
\citep{Haughton1995,Haughton1996,Rahman1993,Bhalotra2008,Kumar2016,Soest2018}.%
\footnote{
Ethnic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.
}
Outside Asia, the evidence is more mixed, with North Africa showing shorter spacing in the 
absence of sons, while a similar pattern does not exist in Sub-Saharan Africa 
\citep{Rossi2015}.

Second, birth spacing can affect children's short and long-term outcomes.
Most of the research finds a negative impact of short spacing on child health, especially 
for very short intervals of 18 months or less \citep{Conde-Agudelo2006,Conde-Agudelo2012}.
There is evidence in countries as diverse as Bangladesh, Brazil, El Salvadore, India,
and Pakistan of worse health outcomes and a higher mortality risk the shorter the birth
interval
\citep{Cleland1984,Curtis1993,Whitworth2002,Bhargava2003,Rutstein2005,Bhalotra2008,Davanzo2008,Maitra2008,Makepeace2008,Gribble2009,Jayachandran2011,Saha2013,Jayachandran2017a,Ghosh2018}.
Most of the evidence on longer-term outcomes come from Western countries.
In the US, for example, a longer interval between births increases the older, but not the
younger, sibling's test scores \citep{Buckles2012}.
Furthermore, close spacing appears to increase the risk of dropping out of
high school and lower the probability of attending a post-secondary school in both the
US and Sweden \citep{Powell1993,Pettersson-Lidbom2009}.%
\footnote{
However, recent results from Sweden show no relationship between birth spacing, even if 
very short, and outcomes such as years of education completed, earnings, and unemployment 
\citep{Barclay2017}.
}

Third, the mother may also be affected by the length of the birth intervals, although 
reviews of the literature show mixed results on mother's anthropometric status 
\citep{Dewey2007,Conde-Agudelo2012}.
For India, there is, however, evidence that women with first-born girls are more likely to 
have anemia, possibly because short birth spacing can resulting in maternal depletion 
\citep{Milazzo2018}.
In Western countries, a longer interval between births has positive effects on long-term 
labor market outcomes, such as participation and income \citep{Gough2017,Karimi2014}.

Finally, we know less about what determines spacing behavior in developing countries than 
in developed countries, and with declining fertility and increasing numbers of women 
entering the labor force, understanding how couples make timing decisions is necessary 
for the design of suitable policies \citep{Portner2018}.%
\footnote{
It is clear that birth spacing does respond to policies in both developed and 
developing countries \citep{Pettersson-Lidbom2009,Todd2012,Meckel2015,Ghosh2018}.
}
Even the effects of access to contraceptive and declining fertility on spacing are unclear.
On the one hand, access to contraceptives allows women to avoid too short spacing between 
birth.
On the other hand, increased reliability of access and effectiveness of contraceptives can 
lead to shorter intervals between births if women used to have longer spacing to avoid
having too many children by accident \citep{Keyfitz1971,Heckman1976}.
These two counteracting effects may explain why better-educated women have shorter spacing 
than less educated women in some instances but not in others and why finding statistically 
significant effects of contraception use on birth spacing is difficult 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Yeakey2009,Kim2010,Soest2018}.%
\footnote{
In Matlab, the effect of son preference on birth spacing is stronger in areas with better 
access to family planning \citep{Rahman1993}.
Other evidence, however, points to families ability to time births, even in the absence
of modern contraceptives \citep{Jayachandran2011,Alam2018}.
}

% [Hypotheses / questions]
Despite the increasing use of sex selection, the critical impacts of spacing on outcomes 
for both mother and children, and the extensive use of spacing to measure son preference, 
there has so far been no systematic examination of what effect access to sex selection 
have on spacing.
I address three questions to remedy this shortcoming in the literature.

@@

Even as the level of female education has increased, the female labor force participation 
in both urban and rural areas has stagnated or decreased
\citep{Klasen2015,Afridi2018,Bhargava2018,Chatterjee2018, Bhargava2019}.
A decline in female labor force participation at the beginning of development is
consistent with the hypothesis of a U-shaped labor force participation as a country
develops \citep{Goldin1994}.
India's female labor force participation is, however, lower than most other
countries and more in line with countries in the Middle East and North Africa, and
does not yet show any signs of increasing
\citep{Klasen2015,Chatterjee2018}.

>>>>

% Demography abstract
\noindent 
Strong son preference is typically associated with shorter birth spacing in the absence of 
sons, but access to sex selection has the potential to reverse this pattern because each 
abortion extends spacing by six to twelve months. 
I apply an empirical method that simultaneously accounts for how sex selection 
increases both birth spacing and the likelihood of a son. 
Using the four rounds of India's National Family and Health Surveys, I show that birth 
intervals increased substantially over the last four decades, with the most significant
increases among women most likely to use sex selection.
Specifically, well-educated women with no boys now have significantly longer birth
intervals, and more male-biased sex ratios, than similar women with boys. 
Women with no education still follow the standard pattern of short average spacing when 
they have girls and little evidence of sex selection, with medium-educated women showing 
mixed results.
Furthermore, the traditional pattern of longer average spacing with sons appears to arise 
not from fewer very short birth intervals, but rather from the upper end of the distribution.

@@

India has changed significantly over the last three decades:
the economy has grown substantially,
education has increased for both males and females,
and the total fertility rate has fallen to the point where, at 2.2, it is just
above replacement 
\citep{Bosworth2008,Guilmoto2013,Dharmalingam2014,International-Institute-for-Population-Sciences-IIPS2017}.
However, access to prenatal sex determination also expanded rapidly starting in the mid-1980s,
and, combined with a continued strong son preference, this dramatically increased the 
males-to-females ratio at birth 
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2009a,Guilmoto2012,
Bongaarts2013,Portner2015b,Jayachandran2017}.

What has not previously been examined is the impact of these changes on birth spacing.
Economic growth, increasing female education, and falling fertility has been
shown to significantly affect birth spacing in other countries [TK references].
India is, however, fundamentally different in at least one respect and that is 
the substantial use of sex-selective abortions.

The stronger the son preference and the lower the desired fertility, the more likely 
couples are to resort to sex-selective abortions, and each abortion increases the 
interval between births by six to twelve months.%
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Hence, the changes in birth spacing may be much larger in India than what we have
observed in other places as they developed.

Understanding how birth spacing changes is important for three reasons.

First, tempo effects, arising from changes in birth spacing, biases the total fertility 
rate (TFR), in the sense that the TFR becomes a more biased measure of cohort fertility
\citep{Bongaarts1999,Ni-Bhrolchain2011}.
Hence, it is possible that fertility is, in fact, substantially higher than what
is generally accepted based on the TFR.

Second, longer birth spacing is associated with lower mortality and 
morbidity risk \citep{Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.
In India, the duration to the next birth has traditionally been shorter with fewer sons, 
which likely contributed to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhargava2003,Rutstein2005,Bhalotra2008,Maitra2008,Makepeace2008,
Jayachandran2011,Kumar2016,Jayachandran2017a,Ghosh2018}.
Longer spacing between births then potentially lowers mortality, although there may
be a counteracting effect of multiple abortions.

Finally, birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals the fewer sons a family has 
\citep{ben-porath76b,Leung1988}.
Instead, longer birth intervals in the absence of sons, which previously would be 
taken as an indicator of lower son preference, may now instead arise because couples 
with a strong son preference use sex selection.

Here, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
I apply an empirical method that directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son. 
The method can be used to analyze both situations with and without prenatal
sex selection.
I apply the method to birth histories of Hindu women, using data from the four
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016. 
I address three main questions.

% [Hypotheses / questions]

@@

Even as the level of female education has increased, the female labor force participation 
in both urban and rural areas has stagnated or decreased
\citep{Klasen2015,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
A decline in female labor force participation at the beginning of development is
consistent with the hypothesis of a U-shaped labor force participation as a country
develops \citep{Goldin1994}.
India's female labor force participation is, however, lower than most other
countries and more in line with countries in the Middle East and North Africa, and
does not yet show any signs of increasing
\citep{Klasen2015,Chatterjee2018}.
